---
title: "A threasure trove of images"
author: "José R. Ferrer-Paris"
date: "2024-05-18"
categories: [how2, flickr, API, R code]
---

So, I think I need some photos in my website, and I have a Flickr account, and I use R. 

There should be a library that... Oh yes! found it!

<https://koki25ando.github.io/FlickrAPI/>

```{r}
#| eval: false
install.packages("FlickrAPI")
```

Now I need a [Flickr API key](https://www.flickr.com/services/developer/). Once I have the key I can save it as an environment variable and call it from R:

```{r}
library(FlickrAPI)
setFlickrAPIKey(api_key = Sys.getenv("FLICKR_API_KEY"))
```

I can query up to 1000 photos from one user using the `getPhotos` function. But I figured out that I could use a `foreach` loop to run and combine multiple queries.

```{r}
library(foreach)
library(dplyr)

photos <- foreach(
  the_user=c("jferrer", "jferrer", "199798864@N08"), 
  the_page=c(1,2,1),
  .combine = "bind_rows") %do% {
  getPhotos(
    user_id = the_user, 
    img_size="m", 
    extras = c("description", "owner_name", "url_m"), 
    per_page=1000, 
    page=the_page)
}

dim(photos)
```

Now I have a collection of photos in R and I can select one by title or any other criteria:

```{r}
selected_photo <- slice(photos, grep("N/W", title))
```

And combine the R and markdown magic to show the photo in this document:

```{r}
#| results: asis
photo_md <- sprintf(
  "![**%s** by %s@flickr](%s){.preview-image .lightbox}",
  selected_photo$title,
  selected_photo$ownername,
  selected_photo$url_m
)

cat(photo_md)
```

If I want to save the information for future use in my blog, I need first to create a folder to hold the file, and then save the R object into an RDS file. Like this:

```{r}
here::i_am("posts/foto-collection.qmd")
data_dir <- here::here("Rdata")
if (!dir.exists(data_dir))
  dir.create(data_dir)
file_name <- here::here(data_dir, "flickr-photos.rds")
saveRDS(file = file_name, photos)
```

