---
title: "iNaturalist: my observations from around the world"
author: "José R. Ferrer-Paris"
date: "2024-05-18"
categories: [how2, iNaturalist, API, R code]
---

It might be nice to share some of my iNaturalist observations in this blog. So I decided to create a backup of observations to quickly add them to my posts.

To get the data I first install the `rinat` package:

```{r}
#| eval: false
install.packages("rinat")
```

Then I can download the observations from iNaturalist:

```{r}
#| eval: true
library(rinat)
library(dplyr)
user_obs <- get_inat_obs_user("NeoMapas",maxresults = 5000)
```


```{r}
glimpse(user_obs)

```

Now I have a collection of photos in R and I can select one by title or any other criteria:

```{r}
selected_photo <- slice(user_obs, 500)
```

And combine the R and markdown magic to show the photo in this document:

```{r}
#| results: asis
photo_md <- sprintf(
  "![*%s* observed in %s by %s@iNaturalist](%s){.preview-image .lightbox}",
  selected_photo$species_guess,
  selected_photo$place_guess,
  selected_photo$user_login,
  selected_photo$image_url
)

cat(photo_md)
```

If I want to save the information for future use in my blog, I need first to create a folder to hold the file, and then save the R object into an RDS file. Like this:


```{r}

here::i_am("posts/foto-collection.qmd")
data_dir <- here::here("Rdata")
file_name <- here::here("Rdata","iNaturalist-obs-NeoMapas.rds")
saveRDS(file=file_name, user_obs)

```

