---
title: "Welcome To My World"
author: "José R. Ferrer-Paris"
date: "2024-05-17"
date-updated: "2024-06-11"
categories: [how2]
engine: knitr
---

This is the typical `Hello World!` post in my new and wonderful blog. Welcome!

So this is basically here to test if the whole thing works.

To get started I first installed quarto.^[Well to be honest, it all started with a operating system, then installing lots of packages, and R and Python, and ... but let's keep it short.]

Then, once I decided how to name my blog, I ran:

```sh
quarto create project blog spatial-one
```

Then I started to prepare the content and structure of my initial posts, modified the text in the `qmd` files and the configuration in the `_quarto.yml` and ran 

I started to keep track of changes with git:^[If we do this before rendering or preview of the site, *quarto* will create a `.gitignore` file with the line `/.quarto/`.]

```sh
cd spatial-one
git init
```

I switch between VS-code, Rstudio and the terminal to organise my files, and kept working on this until I got something I wanted to preview:

```sh
quarto preview spatial-one
```

Here some random notes about the process

## icons

For some reason I though these would fontsawesome icons, but the help prompt in VS-code pointed out:

> Name of bootstrap icon (e.g. github, twitter, share). See https://icons.getbootstrap.com/ for a list of available icons

Also, it turns out we can add emojis to the text of a post using this in the front-matter of the file:

``` yaml
---
title: "My Document"
from: markdown+emoji
---
```

This is a handy demonstration of all emojis: <https://gist.github.com/rxaviers/7360908>

## Freeze

The posts directory has a `_metadata.yml` file with the `freeze` option activated by default.

## `_site` folder

In the quarto documentation they recommend to add it to the `.gitignore` file, so as to keep the raw content and code in version control, but the output served in a different workflow (either with quarto.pub or GitHub actions).

## Secrets

There are certain environment variables that need to be specified in order to render the posts in this blog. I created an `_environment.required` file with the list of required variables and empty values.

```{bash filename = "_environment.required"}
#| eval: false
FLICKR_API_KEY=
```

For local development, I defined these variables in the `_environment.local` file and ignore this file in version control by adding `/_*.local` in `.gitignore`. 

## Engine

Some code is running with RStudio / VScode and R , some code with Python and Jupyter.

For python, we [created a virtual environment](https://quarto.org/docs/projects/virtual-environments.html#using-venv) in the project directory.

```bash
python3 -m venv .venv
```

Then we activate the environment

```bash
source .venv/bin/activate    
```

Then we update pip, install other packages with:

```{.sh}
pip install --upgrade pip
pip install jupyterlab
pip install numpy matplotlib spyder owslib 
pip install geopandas rasterio rasterstats affine osmnx
pip install earthpy
pip install h3pandas
pip install pyinaturalist
pip install pygbif
```

To make this environment reproducible, we create a requirements.txt using pip

```bash 
python3 -m pip freeze > requirements.txt
``` 

Then we can reproduce this environment on another machine by creating an empty environment, activating it, and then:

```bash 
python3 -m pip install -r requirements.txt
``` 

In MacOSX we might need to do this to install the certificates:

```bash
open /Applications/Python\ 3.12/Install\ Certificates.command
```

## Photos

### From Flickr

In order to use some external photos in my blog posts I create a table with the basic information from my photo collection in Flickr. I need to add this code-chunk in my post to locate and load the file:

```{r}
#| echo: fenced
#| message: false
here::i_am("posts/welcome.qmd")
data_dir <- here::here("Rdata")
file_name <- here::here(data_dir, "flickr-photos.rds")
photos <- readRDS(file = file_name)
```

And these lines to create the markdown code to include in my blog post (notice the `results` option):

```{r}
#| echo: fenced
#| results: asis
selected_photo <- dplyr::slice(photos, grep("Del-etreando", title))

photo_md <- sprintf(
  "![**%s** by %s@flickr](%s){.preview-image .lightbox}",
  selected_photo$title,
  selected_photo$ownername,
  selected_photo$url_m
)

cat(photo_md)
```

I add the class `preview-image` to select which foto will appear in the listing page of posts.


### From iNaturalist

Since I also have a bunch of observations in iNaturalist, I can use those as another source of images to illustrate my blog. I need to add this code-chunk in my post to locate and load the file:

```{r}
#| echo: fenced
#| message: false
here::i_am("posts/welcome.qmd")
data_dir <- here::here("Rdata")
file_name <- here::here("Rdata","iNaturalist-obs-NeoMapas.rds")
user_obs <- readRDS(file=file_name)
```

And these lines to create the markdown code to include in my blog post (notice the `results` option):

```{r}
#| echo: fenced
#| results: asis
selected_photo <- user_obs |> dplyr::slice_sample(n=1)
photo_md <- sprintf(
  "![*%s* observed in %s by [%s@iNaturalist](%s)](%s){height=150 group=\"my-gallery\"}",
  selected_photo$species_guess,
  selected_photo$place_guess,
  selected_photo$user_login,
  selected_photo$url,
  selected_photo$image_url
)

cat(photo_md)
```


### From other places

I keep some random images in the `thesaurus` folder and use `imagemagick` to make copies for my blog post. For example:


```{bash}
#| eval: false
convert thesaurus/207d9efb-e858-4c4c-81ab-ac056f72c870.jpeg -resize 300x300 posts/wms-in-R/image.jpg 
```

## Publish / Code

### On Github

I created an empty repo in GitHub and then added it as a remote:

```{bash}
#| eval: false
git remote add origin git@github.com:{...}/{...}.git
git branch -M dont-panic
git push -u origin dont-panic
```

Now the source code and git history is available at: <https://github.com/jrfep/spatial-one>

### Comments through Giscus

How to get comments powered by GitHub Discussions and the giscus app?

1. Enable the [Discussions feature](https://docs.github.com/en/discussions/quickstart) in the GitHub repo
2. Install the [giscus app](https://github.com/apps/giscus) in the GitHub account
3. [Enable comments](https://quarto.org/docs/output-formats/html-basics.html#commenting) in the _quarto.yml file:

``` yaml
---
website:
  comments:
    giscus: 
      repo: jrfep/spatial-one
---
```

### Quarto Pub

I am preparing this site for my quarto.pub profile. From the directory where this project is located, I executed the quarto publish command for Quarto Pub:

```{bash}
#| eval: false
quarto publish quarto-pub
```

And then followed [instructions at quarto.org](https://quarto.org/docs/publishing/quarto-pub.html).

Easy!



